Review Paper Draft

### 2020-07-20 main.tex
3.1 - 3.5: adding unpenalized covariates to models and derivations where appropriate.

\anna{We show this in the absence of fixed covariates for the sake of a clear illustration.. $\bX_0 \bbeta_0$ unpenalized things}

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the normalized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 p_j) / \sqrt{p_j (1 - p_j)}$ and $p_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \text{\anna{unpenalized stuff}} + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\Int$ is the intercept, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

\begin{align}
    \label{eqn:k}
    \bK &= \frac{1}{p} \bX \bX^T \nonumber \\
                   &= \frac{1}{p} \bU \bD^2 \bU^T \nonumber \\
                   &= \frac{1}{p} \bU \bD (\bU \bD)^T \nonumber \\
                   &\propto \bR \bR^T,
\end{align}
where $\bD^2$ is an $n \times n$ diagonal matrix containing the $k$ nonzero eigenvalues of $\bK$ (equivalently, the $k$ nonzero squared singular values of $\bX$) on the top-left diagonal, followed by $n - k$ zeroes on the bottom-right diagonal, and $\bU$ is the corresponding $n \times n$ orthonormal matrix of the left eigenvectors of $\bK$ (equivalently, the left singular vectors of $\bX$.) The first $k$ columns of $\bU$ correspond to the nonzero eigenvalues of $\bK$ (squared singular values of $\bX$.) The columns of $\bR$ are therefore proportional to the principal components of $\bX$ or $\bK$.

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bX$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. It has been shown \citep{hoffman2013correcting} that this corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \Int + \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \eqref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\bbeta, \boldsymbol{\alpha}|\bX, \by) = \frac{1}{2n} ||\by - \Int - \bX \bbeta- \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||.||_2$ the $\ell_2$ norm. Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term. Modeling unobserved confounding using fixed effects in this manner thus requires the estimation of $p + k + 1$ parameters.

@@

\anna{We show this in the absence of fixed covariates for the sake of a clear illustration.. $\bX_0 \bbeta_0$ unpenalized things}

LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N\left(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p \right).$$  Typically, $\bZ$ is not observed directly, and a common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}.This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \Int + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$. Rather than assuming anything about $\bgamma$ directly, an equivalent representation lets $\bu=\bX\bgamma$ such that $\bu \sim N(\boldsymbol{0}, \sigmagg \bK)$, and yields the equivalent model
\begin{equation}
    \label{eqn:lmm_reg_u}
    \by = \Int + \bX\bbeta + \bu + \beps.
\end{equation}
Regardless of representation, the LMM assumes $\by \sim N(\Int + \bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \left(\Int +\bX \bbeta\right), \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
\begin{comment}
For ease of notation, we incorporate the intercept into the design matrix and let $\bX' = \begin{bmatrix} \textbf{1} & \bX \end{bmatrix}$ and $\bbeta' = \begin{pmatrix} \mu & \bbeta \end{pmatrix}^T$ so that we can express this more compactly as 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX' \bbeta', \bI).
\end{equation}
\end{comment}
Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$ and $\byT = \bU^T \left(\by - \Int \right)$, where we center the transformed $\by$ to simplify notation and computation.  $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \bD^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of only $p + 3$ parameters.

\anna{In the presence of unpenalized covariates to adjust for like age and sex... equations 7 and 8 and X-tilde become this... penalty term only penalizes beta1 - betas corresponding to SNPs}

@@

Thus, whether modeled as a fixed or random effect, the underlying effect of $\bZ \bgamma$ on phenotype remains the same, but its treatment as a fixed or random effect has important implications in terms of model fitting and estimation. Including principal components as fixed effects requires estimation of a greater number of parameters, which may result in a loss of power \citep{zhang2015principal}. Additionally, determining the appropriate number of principal components to adjust for is non-trivial, and has been the subject of numerous studies \citep{patterson2006population, zhao2018practical}. It has been shown that including too many or too few principal components can result in power loss or increased Type I errors, respectively \citep{zhang2015principal}. Nevertheless, a common practice is to include the ten largest principal components in the analysis \citep{zhao2018practical}, so this is the way we implement PC-Lasso throughout Section~\ref{sec:results}.

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice \citep{Rakitsch2012, bhatnagar2019simultaneous} is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude ``washes out'' bias in $\hat{\bbeta}$ while inflating the variance. To see why this is the case, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\mu = 0$ for ease of notation: 
\begin{align*}
    \mathbb{E}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps)\\
    &\quad\\
    \mathbb{V}(\hat{\bbeta}) &= (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\by)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps)\\
    &=  (\bX^T \bX)^{-1} \bX^T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^T \bX)^{-1}.
\end{align*}
The assumption that environmental effects follow a random distribution, $\bgamma \sim N(\boldsymbol{0}, \sigmagg / p \bI_p)$, yields $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = \sigmagg/p \bI + \sigmaee (\bX^T \bX)^{-1}$. Returning to our earlier example involving confounding due to air quality, randomly drawing a new value of $\bX\bgamma$ for each replication would mean that subpopulation A may be exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from fixed populations A and B? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if we generate new values of $\bX\bgamma$ with each replication. Considering $\bgamma$ as a fixed quantity is more consistent with the usual meaning of environmental confounding. Specifically, in the low-dimensional OLS setting, under the assumption that the quantity $\bgamma$ is fixed, $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^T \bX)^{-1}$, which allows us to quantify bias in our estimates.

@@

Define the projection matrix of $\bX$ by $\boldsymbol{P}_{\bX} = \bX (\bX^T \bX)^{-1} \bX^T$, and let $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denote the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then, $\exists \boldsymbol{\tau, \psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$ , $\boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$, and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$ such that
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta' + \beps'
\end{align*}
where $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat'$ and $\boldsymbol{\psi}$ is absorbed by the error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.

>>>>

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the normalized version of this genotype matrix as $\bX$, where each $X_{ij} = (G_{ij} - 2 p_j) / \sqrt{p_j (1 - p_j)}$ and $p_j$ is the minor allele frequency (MAF) of SNP $j$ \citep{zhang2015principal, price2006principal}. Additionally, we allow $\bX_0$ to represent an $n \times p_0$ matrix of unpenalized covariates. At a minimum, $\bX_0$ will be an $n \times 1$ vector of 1s to represent the intercept, in which case $\bbeta_0 \equiv \beta_0$ is the scalar intercept. However, $\bX_0$ may additionally include covariates, such as age and sex, which are of interest to include in the model without penalty. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of unpenalized covariates, additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \bX_0 \bbeta_0 + \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta_0$ is a $p_0 \times 1$ vector which includes the intercept term and the effects of any unpenalized covariates, $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bgamma$ is a $q \times 1$ vector of environmental effects, $\bZ$ is an $n \times q$ matrix that allocates environmental effects among subjects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$. Often it is assumed that subjects belong to $q \le p$ subpopulations, and $\bZ$ is a corresponding incidence matrix of 0s and 1s. However, as we will discuss, this is not the only possibility.

@@

\begin{align}
    \label{eqn:k}
    \bK &= \frac{1}{p} \bX \bX^\T \nonumber \\
                   &= \frac{1}{p} \bU \bD^2 \bU^\T \nonumber \\
                   &= \frac{1}{p} \bU \bD (\bU \bD)^\T \nonumber \\
                   &\propto \bR \bR^\T,
\end{align}
where $\bD^2$ is an $n \times n$ diagonal matrix containing the $k$ nonzero eigenvalues of $\bK$ (equivalently, the $k$ nonzero squared singular values of $\bX$) on the top-left diagonal, followed by $n - k$ zeroes on the bottom-right diagonal, and $\bU$ is the corresponding $n \times n$ orthonormal matrix of the left eigenvectors of $\bK$ (equivalently, the left singular vectors of $\bX$.) The first $k$ columns of $\bU$ correspond to the nonzero eigenvalues of $\bK$ (squared singular values of $\bX$.) The columns of $\bR$ are therefore proportional to the principal components of $\bX$ or $\bK$.

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bX$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. It has been shown \citep{hoffman2013correcting} that this corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \bX_0\bbeta_0 + \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \eqref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\bbeta, \boldsymbol{\alpha}|\bX, \by) = \frac{1}{2n} ||\by - \bX_0\bbeta_0 - \bX \bbeta- \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, and $||.||_2$ the $\ell_2$ norm. Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term.

@@

LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N\left(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p \right).$$  Typically, $\bZ$ is not observed directly, and a common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}.This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \bX_0\bbeta_0 + \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$. Rather than assuming anything about $\bgamma$ directly, an equivalent representation lets $\bu=\bX\bgamma$ such that $\bu \sim N(\boldsymbol{0}, \sigmagg \bK)$, and yields the equivalent model
\begin{equation}
    \label{eqn:lmm_reg_u}
    \by = \bX_0\bbeta_0 + \bX\bbeta + \bu + \beps.
\end{equation}
We will show LMM-Lasso derivations in the absence of unpenalized covariates for the sake of a clear illustration.

In the absence of unpenalized covariates, the LMM assumes $\by \sim N(\bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\label{eqn:rtY_distn}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}} \bX \bbeta, \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
Using the spectral decomposition of $\bK$ described in equation \eqref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization 
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^\T + \sigmaee \bU \bU^\T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^\T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^\T\\
    &=\bU \bW^2 \bU^\T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^\T \bX$ and $\byT = \bU^\T \by$, where $\bW = \text{diag}\{w_1, ... w_n\}$ and $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \bD^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
In the presence of an intercept and unpenalized covariates, \eqref{eqn:rtY_distn} becomes 
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N\left(\bV^{-\frac{1}{2}} \bX{^\prime} \bbeta{^\prime}, \bI \right)
\end{equation}
where $\bX{^\prime} = \begin{bmatrix} \bX_0 & \bX \end{bmatrix}$ and $\bbeta{^\prime} = \begin{bmatrix} \bbeta_0^\T & \bbeta^\T \end{bmatrix}^\T$. This in turn leads to a new version of \eqref{eqn:lmm_obj}, our objective function,  
\begin{equation}
Q(\bbeta | \bX{^\prime}, \byT, \bW) = || \bW (\byT - \bXT{^\prime} \bbeta{^\prime})||_2^2 + \lambda || \bbeta ||_1
\end{equation}
where $\bXT{^\prime} = \bU^\T \bX{^\prime}$, and only the elements of $\bbeta{^\prime}$ corresponing to SNPs are penalized.

@@

Thus, whether modeled as a fixed or random effect, the underlying effect of $\bZ \bgamma$ on phenotype remains the same, but its treatment as a fixed or random effect has important implications in terms of model fitting and estimation. Modeling unobserved confounding using fixed effects requires estimation of $p_0 + p + k$ parameters, compared to the $p_0 + p + 2$ required of an LMM. This may result in a loss of power \citep{zhang2015principal}. Additionally, determining the appropriate number of principal components to adjust for is non-trivial, and has been the subject of numerous studies \citep{patterson2006population, zhao2018practical}. It has been shown that including too many or too few principal components can result in power loss or increased Type I errors, respectively \citep{zhang2015principal}. Nevertheless, a common practice is to include the ten largest principal components in the analysis \citep{zhao2018practical}. This is the implementation of PC-Lasso throughout Section \ref{sec:results}.

@@

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice \citep{Rakitsch2012, bhatnagar2019simultaneous} is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude ``washes out'' bias in $\hat{\bbeta}$ while inflating the variance. To see why this is the case, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align*}
    \mathbb{E}(\hat{\bbeta}) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by)\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps)\\
    &\quad\\
    \mathbb{V}(\hat{\bbeta}) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by)\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps)\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1}.
\end{align*}
The assumption that environmental effects follow a random distribution, $\bgamma \sim N(\boldsymbol{0}, \sigmagg / p \bI_p)$, yields $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = \sigmagg/p \bI + \sigmaee (\bX^\T \bX)^{-1}$. Returning to our earlier example involving confounding due to air quality, randomly drawing a new value of $\bX\bgamma$ for each replication would mean that subpopulation A may be exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from fixed populations A and B? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if we generate new values of $\bX\bgamma$ with each replication. Considering $\bgamma$ as a fixed quantity is more consistent with the usual meaning of environmental confounding. Specifically, in the low-dimensional OLS setting, under the assumption that the quantity $\bgamma$ is fixed, $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, which allows us to quantify bias in our estimates.

@@

Define the projection matrix of $\bX$ by $\boldsymbol{P}_{\bX} = \bX (\bX^\T \bX)^{-1} \bX^\T$, and let $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denote the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then, $\exists \boldsymbol{\tau, \psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$ , $\boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$, and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$ such that
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta{^*} + \beps{^*}
\end{align*}
where $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat{^\prime}$ and $\boldsymbol{\psi}$ is absorbed by the error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.
